---
title: "Demo_3"
author: "Shuyu Li"
date: "2025-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(bio3d)
```

## Reading multi-model PDB files

To enable read multi-model PDB file set `multi=TRUE`. The structure solved from NMR c

```{r}
# Read multi-model PDB file
pdb <- read.pdb("1dmo", multi=TRUE)
```

```{r}
# The xyz component contains 20 frames
pdb$xyz

```

```{r}
# Select a subset of the protein
ca.inds <- atom.select(pdb, "calpha")

# Access C-alpha coordinates of the first 5 models
#pdb$xyz[1:5, ca.inds$xyz]
```

## Identification of dynamic domains

Function **geostas()** attempts to identify rigid domains in a protein (or nucleic acid) structure based on knoweledge of its structural ensemble (as obtained a from multi-model PDB input file, an ensemble of multiple PDB files (see section below), NMA or MD results). 

The `geostas()` function in **Bio3D** analyzes a structural ensemble and tries to:

-   Cluster atoms (typically Cα atoms in proteins) into **rigid groups**.

<!-- -->

-   These groups are defined based on **how little their relative distances change** across the ensemble.

-   It returns a classification of residues into **rigid domains** and can help **visualize domain motion**.

```{r}
# Domain analysis
gs  <- geostas(pdb)
```

```{r}
# Fit all frames to the 'first' domain
domain.inds <- gs$inds[[1]]

xyz <- pdbfit(pdb, inds=domain.inds)

# write fitted coordinates
write.pdb(pdb, xyz=xyz, chain=gs$atomgrps, file="1dmo_fit-domain1.pdb")

# plot geostas results
plot(gs, contour=FALSE)
```

## Invariant core identification

```{r}
core <- core.find(pdb)

xyz <- pdbfit(pdb, inds=core)

write.pdb(pdb, xyz=xyz, file="1dmo_fit_core.pdb")

```
